\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-09-04 introduction.tex
reply
intro

The EP is a programming challenge that
epitomizes the difficulty of writing type-safe, extensible code.
To define an expression language that can be reused for future extensions,
the programmer faces a fundamental tension \cite{reynolds1975} between
adding new constructors of a data type (e.g., new abstract syntax) and
adding new functions over the data type (e.g., new compiler passes).

@@

In contrast, proof assistants offer few linguistic solutions that
address the EP.
Yet the challenge of writing extensible, type-safe code (and proofs!) is
as real, especially for metatheory mechanization.
As an example, consider \cref{fig:STLC-example}, the simply typed
lambda calculus (STLC) mechanized in Coq.
A programming challenge there is to define distinct extensions of this
STLC formalization to support distinct features,
and then selectively compose these extensions to form new STLC variants.

In Coq, inductive types, as well as functions and theorems over
inductive types, are closed to extension.
So to reuse mechanized metatheories,
the common practice is still to copy and modify them for each extension.
But having to maintain multiple copies is highly non-modular and
antithetical to good software engineering.
%
The programmer could also turn to design patterns~\cite{delaware2011,delaware2013}.
But they require heavy lifting from the programmer to make code
extensible, often leading to verbose, non-idiomatic programming styles.

A question thus arises: can linguistic solutions to the EP, as studied
in the conventional setting, be adapted to metatheory mechanization in
a proof assistant?

At the core of many linguistic solutions to the EP is \emph{inheritance}.
Inheritance is sometimes interpreted narrowly as a subclass'
inheriting methods and instance variables from its superclass.
But the language-theoretic essence of inheritance is more general:
it is a linguistic approach to incrementally modifying
\emph{recursive, record-like definitions}~\cite{cook1990inheritance}.

Language mechanisms including
mixins~\cite{mixin-1990},
virtual classes~\cite{virtualclasses-1989,vc-calculus-2006},
virtual types~\cite{thorup97} and associated types~\cite{ckj05},
extensible cases~\cite{bac2006},
and so on, are all based on the essential idea of inheritance.
%
In particular, when a language mechanism enables inheritance over
recursive definitions containing related types and terms,
it is said to support \emph{family polymorphism}~\cite{ernst2001family}:
types and terms are polymorphic to a family they are nested within.

\paragraph{Main ideas.}

We contribute a language design that integrates family polymorphism into
a proof assistant.
Because code and proofs are polymorphic to a family they are nested
within, they can be inherited and reused by a derived family.
%
Family polymorphism enables extensible metatheory mechanization.
As \cref{fig:?} shows,
different extensions to STLC can all inherit from the base STLC family:
they reuse mechanized metatheories from syntax to the type safety
theorem, only adding new constructors to inductive types
and new cases to the induction proofs as needed by an extension.

@@

\paragraph{Contributions} To our knowledge, we offer the first
language-based solution that enables extensible metatheory mechanization
in a higher-order, dependent type theory with inductive types.

@@

\item We show how to integrate family polymorphism into a proof
assistant (\cref{sec:lang-design}).
The language design reconciles the expressiveness enabled by
family polymorphism with the exactness of a proof assistant,
while retaining an idiomatic programming style.

@@

\item We capture the new language mechanism formally by extending
Martin–Löf type theory with family polymorphism and extensible inductive
types (\cref{sec:metatheory}).
We derive strong meta\-theoretical results including on consistency and
canonicity.
We also formalize the translation implemented by the Coq plugin.

>>>>

The EP is a programming challenge that
epitomizes the difficulty of writing type-safe, extensible code.
To define an expression language that can be reused for future extensions,
the programmer faces a fundamental tension \cite{reynolds1975} between
adding new constructors to a data type (e.g., new abstract syntax) and
adding new functions over the data type (e.g., new compiler passes).

@@

In contrast, proof assistants offer few linguistic solutions that
address the EP.
Yet, the challenge of writing extensible, type-safe code is
as real, especially for metatheory mechanization.
The programmer faces a tension between adding new constructors to an inductive data type
(e.g., new abstract syntax) and adding new functions and theorems over
the data type (e.g., new meta\-theoretical results).

In the Coq proof assistant~\cite{coq}, inductive types, as well as functions and
theorems over inductive types, are closed to extension.
So to reuse mechanized metatheories,
the common practice is still to copy code and proofs and then modify them in each extension.
But having to maintain multiple copies is highly non-modular and
antithetical to good software engineering.
%
The programmer could also turn to design patterns~\cite{delaware2011,delaware2013}.
But they tend to require heavy lifting from the programmer to make code
extensible, often leading to verbose, non-idiomatic programming styles.

We answer the following question:
can linguistic features designed to solve the EP (in
conventional OO and functional languages)
be adapted to metatheory mechanization in a proof assistant?

At the core of many linguistic solutions to the EP is \emph{inheritance}.
Inheritance is sometimes interpreted narrowly as a subclass'
inheriting methods and instance variables from its superclass.
But the language-theoretic essence of inheritance is more general:
it is a linguistic approach to \emph{incrementally modifying
recursive definitions}~\cite{cook1990inheritance}.

Language mechanisms including
mixins~\cite{mixin-1990},
virtual classes~\cite{virtualclasses-1989,vc-calculus-2006},
virtual types~\cite{thorup97} and associated types~\cite{ckj05},
extensible cases~\cite{bac2006},
and so on, are all based on this essential idea of inheritance.
%
In particular, when a language mechanism enables inheritance over
recursive definitions containing related types and terms,
it is said to support \emph{family polymorphism}~\cite{ernst2001family}:
types and terms are polymorphic to a family they are nested within.

\begingroup

\paragraph{Contributions}

We contribute a language design that integrates family polymorphism into
a proof assistant.
Because code and proofs are polymorphic to a family they are nested
within, they can be

\begin{wrapfigure}[6]{r}{.215\textwidth}
\includegraphics[scale=.48]{graphics/stlc-intro.pdf}
\end{wrapfigure}

\noindent
inherited and reused by a derived family—%
family polymorphism enables extensible metatheory mechanization.

%As an example, consider \cref{fig:STLC-example}, the simply typed
%lambda calculus (STLC) mechanized in Coq.
%A programming challenge there is to define distinct extensions of this
%STLC formalization to support distinct features,
%and then selectively compose these extensions to form new STLC variants.
As an example, the diagram to the right depicts an extensible
mechanization of the simply typed λ-calculus (STLC), using family
polymorphism.
Different extensions to STLC can all inherit from the base STLC family:
they reuse mechanized metatheories, from syntax to the type safety
theorem, only adding new constructors to inductive types
and new cases to the induction proofs and recursive functions as needed
by an extension.

@@

\endgroup

Specifically, we make the following contributions.

@@

\item We present the first language design that enables extensible
metatheory mechanization in a higher-order, dependent type theory with
inductive types (\cref{sec:lang-design}).
The language design reconciles the expressiveness enabled by
family polymorphism with the exactness of a proof assistant,
while retaining an idiomatic programming style.

@@

\item We capture the essence of the new language mechanism formally by extending
Martin–Löf type theory with family polymorphism and extensible inductive
types (\cref{sec:metatheory}).
We derive strong meta\-theoretical results including consistency and
canonicity.
We also formalize the translation implemented by the Coq plugin.
